Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-25 mr-review.Rmd
citation links added

---
title: "Mendelian randomization for causal inference in Genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

Another approach to integrating multiple SNP instruments into a MR framework is
through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
calculate an unweighted score as the number of exposure-increasing alleles in 
the subject's genotypes. They also calculate a weighted allele score by using 
the exposure effects as weights. For example, a subject with $g_k$ copies of 
exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{ik}$ 
and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.

@@

library(dagitty)
library(ggdag)
library(ggplot2)

@@

@brown_welch-weighted_2021

>>>>

---
title: "Mendelian randomization for causal inference in Genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

Another approach to integrating multiple SNP instruments into a MR framework is
through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
calculate an unweighted score as the number of exposure-increasing alleles in 
the subject's genotypes. They also calculate a weighted allele score by using 
the exposure effects as weights. For example, a subject with $g_k$ copies of 
exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.

@@

library(dagitty)
library(ggplot2)

@@

\begin{eqnarray}
U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
  (\#eq:wwer)
\end{eqnarray}

@oconnor_distinguishing_2018

In efforts to develop MR methods that are robust to correlated horizontal pleiotropy, 
@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
accommodates both uni-directional and bi-directional effects between exposure and outcome. 
@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$. $Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which . Note that the model is agnostic with regards to which is the 
outcome and which the exposure, so we follow the authors' notation here.

In Equations \@ref(eq:wwer),

we use A and B to denote the observed traits in
the study, and U to denote the unobserved genetic confounder.
SNPs X affect each of A, B, and U with probabilities q, r, and s
and effect sizes bA ; bB ; and bU sampled from a normal distribution
with variances s2A ; s2B ; and s2U , respectively. The probability of ef-
fect and variance of the sampling distribution combine to deter-
mine the genetic architecture of each trait independently of the
others. Finally, h and n specify the effect of the hidden confounder
U on A and B, while g and d model the causal effect of A on B and B
on A, respectively. Under this model, the phenotype values are
given by
U ¼ XbU +ZU þ εU ;(Equation 1)
A ¼ Un þ Bd þ XbA +ZA þ εA ;(Equation 2)
B ¼ Uh þ Ag þ XbB +ZB þ εB ;(Equation 3)
where Z’s represent indicator variables that the SNP affects that
trait, sampled as indicated above, + indicates vector element-
wise (Hadamard) multiplication, and bolding represents matrices.
The recursive nature of this model makes sampling from it non-
trivial. In the supplemental methods, we describe how to simulate
from this model and how to parameterize the model in terms of
the heritability of each phenotype rather than the variance of
the effect size distribution. We also explicitly describe how to set
the parameters to mimic the models considered in O’Connor et
al.5 and Morrison et al.3
